Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2020-11-20 Manuscript/HI_Model_ms.Rmd
updated manuscript files

## Abstract
The State of Hawai`i passed legislation to be carbon neutral by 2045, a goal that will partly depends on carbon sequestration by terrestrial ecosystems. However, there is considerable uncertainty surrounding the future direction and magnitude of the land carbon sink in the Hawaiian Islands. We used simulation modeling to assess how projected future changes in climate and land use will influence ecosystem carbon balance in the Hawaiian Islands under all combinations of two radiative forcing scenarios (RCPs 4.5 and 8.5) and two land-use scenarios (low and high) over a 90-year timespan from 2010-2100. Collectively, ecoystems of the Hawaiian Islands acted as a net carbon sink under low radiative forcing throughout the 90-year simulation period, but carbon sink strength was greatest under the combination of low radiative forcing and low land-use change. In contrast, Hawaiian terrestrial ecosystems transitioned from a net sink to a net source of CO~2~ to the atmosphere under high radiative forcing, with high land-use accelerating this transition and exacerbating net carbon loss. A sensitivity test of the CO~2~ fertilization effect on plant productivity revealed it to be a major source of uncertainty in projections of ecosystem carbon balance. Reconciling this uncertainty in how net photosynthesis will respond to rising atmospheric CO~2~ will be essential to realistically constrain simulation models used to evaluate the effectiveness of ecosystem‐based climate mitigation strategies.

## Introduction
Terrestrial ecosystems are a major sink for atmospheric carbon dioxide (CO~2~), removing ~30% of human emissions on an annual basis [@friedlingstein_global_2019].

@@

## Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands under all combinations of two land-use scenarios (low and high) and two radiative forcing scenarios (RCP 4.5 and RCP 8.5). We also developed a separate set of scenarios to test model sensitivity to different levels of a CO~2~ fertilization effect (CFE). The landscape change portion of LUCAS is a state-and-transition model that applies a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transitions [@daniel_state-and-transition_2016]. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying stock level rates of change over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawai`i LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

@@

### *CO~2~ fertilization effect*
Increasing atmospheric CO~2~ concentrations stimulate leaf-level photosynthesis, potentially increasing NPP as well [@Walker_integrating_2020]. However, the magnitude and persistence of this effect is highly uncertain, particularly across a range of climatic conditions and over long time spans [@Walker_integrating_2020]. Following @sleeter_effects_2019, we developed a separate set of scenarios designed to test the sensitivity of LUCAS model ecosystem carbon balance projections to  different rates of a CO~2~ fertilization effect (CFE). We incorporated a CFE multiplier for NPP that represented the percent increase in NPP for every 100 ppm increase in atmospheric CO~2~ concentration under the high land use and high radiative forcing (RCP 8.5) scenario. We tested five CFE levels ranging from 5% to 15%, which is within the range of CFEs observed in free air CO~2~ enrichment (FACE) experiments. For all levels, we assumed CFEs reached saturation at an atmospheric CO~2~ concentration of 600 ppm, with no further stimulation of NPP despite a continued increase in CO~2~ concentration to 930 ppm by 2100. This 600ppm threshold generally coincides with the upper limit from FACE experiments and is reached by the year 2060 under RCP 8.5.

@@

### *Ecosystem carbon balance*
Net biome productivity (NBP) averaged approximately 0.6 Tg C y^-1^ at the start of the simulation period and declined over time in all four scenarios (figure 4). On average, terrestrial ecosystems of the seven main Hawaiian Islands collectively acted as a net carbon sink throughout the simulation period under the RCP 4.5 radiative forcing scenario, but carbon sink strength was ~40% lower in the high land-use scenario compared to the low land-use scenario by the end of the simulation period (figure 4). In contrast, ecosystems of the Hawaiian Islands acted as a net carbon source to the atmosphere toward the latter half of the simulation period under RCP 8.5, with the transition from sink to source occuring 15 years earlier on average in the high land-use scenario than in the low land-use scenario (figure 4). In addition, the high land-use scenario under RCP 8.5 represented a ~40% larger net source of carbon to the atmosphere by the year 2100 than the low-land use scenario under the same radiative forcing. Over the entire simulation period, both global emissions reduction and avoided land conversion resulted in substantial increases in cumulative NBP (figure 5). However, switching from RCP 8.5 to RCP 4.5 increased cumulative NBP in the Hawaiian Islands more than twice as much as reducing emissions from local land-use change and wildfire disturbance (figure 5). Switching from RCP 8.5 to RCP 4.5 under the low land-use scenario yielded the greatest cumulative increase in NBP, resulting in a median gain of 26.5 Tg of carbon over the entire 90-year simulation period.

@@

This study was funded by the U.S. Geological Survey Biological Carbon Sequestration Program. Thanks to Nicholas Koch of Forest Solutions, Inc. for information on eucalyptus harvesting in Hawaii and to Leonardo Frid and Colin Daniel of ApexRMS for assistance with SyncroSim software. Thanks also to Christian Giardina and Zhiliang Zhu for making this project possible. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

>>>>

## Abstract
The State of Hawai`i passed legislation to be carbon neutral by 2045, a goal that will partly depends on carbon sequestration by terrestrial ecosystems. However, there is considerable uncertainty surrounding the future direction and magnitude of the land carbon sink in the Hawaiian Islands. We used simulation modeling to assess how projected future changes in climate and land use will influence ecosystem carbon balance in the Hawaiian Islands under all combinations of two radiative forcing scenarios (RCPs 4.5 and 8.5) and two land-use scenarios (low and high) over a 90-year timespan from 2010-2100. Collectively, terrestrial ecosystems of the Hawaiian Islands acted as a net carbon sink under low radiative forcing (RCP 4.5) for the entire 90-year simulation period, with low land-use change further enhancing carbon sink strength. In contrast, Hawaiian terrestrial ecosystems transitioned from a net sink to a net source of CO~2~ to the atmosphere under high radiative forcing (RCP 8.5), with high land-use accelerating this transition and exacerbating net carbon loss. A sensitivity test of the CO~2~ fertilization effect on plant productivity revealed it to be a major source of uncertainty in projections of ecosystem carbon balance. Reconciling this uncertainty in how net photosynthesis will respond to rising atmospheric CO~2~ will be essential to realistically constrain simulation models used to evaluate the effectiveness of ecosystem‐based climate mitigation strategies.

## Introduction
Terrestrial ecosystems are a major sink for atmospheric carbon dioxide (CO~2~), removing ~30% of human emissions on an annual basis [@friedlingstein_global_2019]. However, the strength of the land carbon sink remains highly uncertain because of its sensitivity to interannual variability in climate, land use, and land cover change [@ahlstrom_dominant_2015; @prestele_current_2017]. Accounting for climate and land-use variability has proven so difficult that annual estimates of terrestrial ecosystem carbon sequestration are often estimated indirectly by subtracting annual atmospheric and ocean carbon sink estimates from annual human CO~2~ emissions [@le_quere_global_2013; @friedlingstein_global_2019].

@@

## Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands under all combinations of two land-use scenarios (low and high) and two radiative forcing scenarios (RCP 4.5 and RCP 8.5). We also developed a separate set of scenarios to test model sensitivity to different levels of a CO~2~ fertilization effect (CFE). The landscape change portion of LUCAS is a state-and-transition model that applies a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transitions [@daniel_state-and-transition_2016]. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawai`i LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

@@

### *CO~2~ fertilization effect*
Increasing atmospheric CO~2~ concentrations stimulate leaf-level photosynthesis, potentially increasing NPP as well [@Walker_integrating_2020]. However, the magnitude and persistence of this effect is highly uncertain, particularly across a range of climatic conditions and over long time spans [@Walker_integrating_2020]. Following @sleeter_effects_2019, we developed a separate set of scenarios designed to test the sensitivity of LUCAS model projections of ecosystem carbon balance to different rates of a CO~2~ fertilization effect (CFE). We incorporated a CFE multiplier for NPP that represented the percent increase in NPP for every 100 ppm increase in atmospheric CO~2~ concentration under the high land use and high radiative forcing (RCP 8.5) scenario. We tested five CFE levels ranging from 5% to 15%, which is within the range of CFEs observed in free air CO~2~ enrichment (FACE) experiments. For all levels, we assumed CFEs reached saturation at an atmospheric CO~2~ concentration of 600 ppm, with no further stimulation of NPP despite a continued increase in CO~2~ concentration to 930 ppm by 2100. This 600ppm threshold generally coincides with the upper limit from FACE experiments and is reached by the year 2060 under RCP 8.5.

@@

### *Ecosystem carbon balance*
Net biome productivity (NBP) averaged approximately 0.6 Tg C y^-1^ at the start of the simulation period and declined over time in all four scenarios (figure 4). On average, terrestrial ecosystems of the seven main Hawaiian Islands collectively acted as a net carbon sink throughout the simulation period under the RCP 4.5 radiative forcing scenario, but carbon sink strength was ~40% lower in the high land-use scenario compared to the low land-use scenario by the end of the simulation period (figure 4). In contrast, ecosystems of the Hawaiian Islands acted as a net carbon source to the atmosphere toward the latter half of the simulation period under RCP 8.5, with the transition from sink to source occuring 15 years earlier on average in the high land-use scenario than in the low land-use scenario (figure 4). The high land-use scenario under RCP 8.5 represented a ~40% larger net source of carbon to the atmosphere by the year 2100 than the low-land use scenario under the same radiative forcing. Over the entire simulation period, both global emissions reductions and local avoided land conversion resulted in substantial increases in cumulative NBP (figure 5). However, switching from RCP 8.5 to RCP 4.5 increased cumulative NBP in the Hawaiian Islands more than twice as much as reducing emissions from local land-use change and wildfire disturbance (figure 5). Switching from RCP 8.5 to RCP 4.5 under the low land-use scenario yielded the greatest cumulative increase in NBP, resulting in a median gain of 26.5 Tg of carbon over the entire 90-year simulation period.

@@

This study was funded by the U.S. Geological Survey Biological Carbon Sequestration Program. Thanks to Leonardo Frid and Colin Daniel of ApexRMS for assistance with SyncroSim software and to Nicholas Koch of Forest Solutions, Inc. for information on eucalyptus harvesting in Hawaii. Thanks also to Christian Giardina and Zhiliang Zhu for making this project possible. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.
